---
title: "Algorithmic design and acknowledgements"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Algorithmic design and acknowledgements}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`fastEmbedR` is built around a KNN-first design. Nearest-neighbour search is
handled by the companion `faissR` package; `fastEmbedR` consumes those
neighbours to run UMAP and openTSNE-style t-SNE embeddings.

This vignette explains what is implemented, what was removed, and which papers
or software projects inspired the current code.

## Package Boundaries

The current project is split into two packages.

| Package | Main responsibility |
| --- | --- |
| `faissR` | FAISS/cuVS nearest neighbours, candidate KNN, KNN graphs, KNN prediction, and k-means. |
| `fastEmbedR` | UMAP, openTSNE-style t-SNE, landmark transforms, and quality metrics. |

Use `faissR::nn()` directly so the same KNN object can be shared across UMAP
and openTSNE runs.

```{r, eval = FALSE}
library(fastEmbedR)

knn <- faissR::nn(x, k = 50, backend = "faiss_ivf_flat", n_threads = 4)
y_umap <- umap_knn(knn, backend = "cpu", seed = 1)
y_tsne <- opentsne_knn(knn, init_data = x, backend = "cpu", seed = 1)
```

The practical advantage is that KNN time and embedding time are separated.
This is important for fair comparisons with `uwot`, `Rtsne`, FIt-SNE, and GPU
implementations.

## UMAP

### CPU UMAP

The CPU UMAP path is native C++ and operates directly from a supplied KNN
object. The implementation keeps the standard UMAP structure:

1. estimate smooth KNN bandwidths;
2. compute local connectivity offsets;
3. build fuzzy graph weights;
4. symmetrize the graph;
5. precompute `epochs_per_sample`;
6. optimize the two-dimensional layout with UMAP's attractive and repulsive
   stochastic objective.

The main improvements are engineering improvements, not changes to the
mathematics:

- compact graph buffers;
- CSR/COO-style graph storage;
- no dense graph intermediates;
- fewer avoidable copies between R and C++;
- `float` internal weights where safe;
- parallel row-wise work where deterministic enough;
- native sampled-edge optimization with negative sampling and learning-rate decay.

### `graph_mode`

UMAP has two graph modes:

```{r, eval = FALSE}
y_binary <- umap_knn(knn, graph_mode = "binary")
y_fuzzy <- umap_knn(knn, graph_mode = "fuzzy")
```

`graph_mode = "fuzzy"` uses the standard UMAP fuzzy graph. `graph_mode =
"binary"` uses binary neighbour edges and can increase visible separation on
some datasets. It is not hidden as an internal trick: benchmark output records
which graph mode was used.

### Metal UMAP

The Metal UMAP backend is implemented in Objective-C++/Metal. It does not call
Python, Torch, MLX, or `reticulate`.

The public Metal optimizer is the visually validated `atomic_inplace` path.
Earlier scheduled, deterministic, and hybrid experiments were removed from the
public API because they made the package harder to reason about and produced
less reliable plots.

```{r, eval = FALSE}
y_metal <- umap_knn(knn, backend = "metal", seed = 1)
```

If Metal is not available, this request fails clearly. It is not silently run
on CPU and reported as Metal.

### CUDA UMAP

The CUDA UMAP backend uses native CUDA kernels when `fastEmbedR` is built with
CUDA support. The cleaned public path is the pure atomic CUDA optimizer. Older
CUDA hybrid and deterministic variants were removed.

CUDA KNN is not implemented in `fastEmbedR`; it comes from `faissR`. The
one-call embedding API uses FAISS GPU IVF-Flat on CUDA. KNN-first workflows can
use other explicit `faissR::nn()` backends for benchmarking.

```{r, eval = FALSE}
knn_cuda <- faissR::nn(x, k = 50, backend = "faiss_gpu_ivf_flat", n_threads = 4)
y_cuda <- umap_knn(knn_cuda, backend = "cuda", seed = 1)
```

## openTSNE-Style t-SNE

`opentsne_knn()` implements an openTSNE-style t-SNE optimizer from precomputed
KNN matrices.

The high-level steps are:

1. convert neighbour distances to perplexity affinities;
2. symmetrize sparse probabilities;
3. initialize the layout, usually with PCA;
4. run early exaggeration;
5. run the normal optimization phase;
6. return the layout plus timing/backend metadata.

```{r, eval = FALSE}
y_tsne <- opentsne_knn(
  knn,
  init_data = x,
  perplexity = 15,
  backend = "cpu",
  seed = 1
)
```

### PCA Initialization

PCA initialization is explicit. Provide `Y_init` or `init_data` when you want
it. For t-SNE this initialization is centered and scaled to the small scale
expected by t-SNE:

```{r, eval = FALSE}
Y_init <- opentsne_pca_init(x)
y_tsne <- opentsne_knn(knn, Y_init = Y_init, perplexity = 15)
```

In MNIST 70k tests, PCA initialization reduced cluster splitting without
changing the objective.

### CPU FFT-Grid Path

The CPU openTSNE large-data path uses a FIt-SNE-style grid approximation for
the repulsive force. The attractive force remains sparse and KNN-based. The
old Barnes-Hut path was removed from the public package surface because the
FFT-grid path was the useful large-data standard in the benchmarks.

### Metal openTSNE

The Metal openTSNE backend is package-native Objective-C++/Metal:

- grid scatter kernels;
- native Metal FFT-grid convolution;
- grid gather kernels;
- sparse attractive force kernels;
- gain, momentum, update, and centering kernels.

MPSGraph FFT was tested as a diagnostic alternative. It remains diagnostic-only
because it did not provide enough improvement over the package-native FFT path.

```{r, eval = FALSE}
y_tsne_metal <- opentsne_knn(
  knn,
  Y_init = Y_init,
  perplexity = 15,
  backend = "metal",
  seed = 1
)
```

### CUDA openTSNE

The CUDA openTSNE backend uses native CUDA kernels and cuFFT when the package
is built with CUDA support:

- device-side grid scatter/gather;
- cuFFT convolution;
- sparse attractive forces;
- optimizer updates on device.

The CUDA path is designed to match the same openTSNE/FIt-SNE formulation used
by the CPU and Metal paths.

## Landmarking

Landmarking is an explicit approximation. It is useful for very large data or
interactive workflows, but it should be labelled as landmarking in benchmarks.

```{r, eval = FALSE}
fit_landmark <- landmark_tsne(
  x,
  landmarks = 0.5,
  n_neighbors = 50,
  perplexity = 15,
  seed = 1
)
```

The package separates:

- embedding time for landmarks;
- projection or transform time for non-landmark points;
- quality metrics for the final full layout.

## Autotuning

The default API asks for few parameters, but the code records what it chose.

openTSNE uses opt-SNE-style rules:

- automatic learning rate is `n / early_exaggeration`;
- iteration defaults are chosen by a native helper;
- CPU/small exact runs can use KLD-based stopping;
- large FFT/GPU runs do not hide an expensive CPU KLD monitor.

UMAP uses the KNN distance profile to choose internal defaults such as
initialization effort and optimizer settings. If a KNN object is supplied, the
KNN graph itself is not silently changed.

## Backend Honesty

The backend rule is simple:

- `backend = "metal"` must run native Metal code or fail;
- `backend = "cuda"` must run native CUDA/cuFFT code or fail;
- CPU fallback is never reported as GPU work.

Use `faissR::backend_info()` to inspect the current KNN build:

```{r, eval = FALSE}
faissR::backend_info()
```

## References

Key references are listed in AACR journal style.

1. van der Maaten L, Hinton G. Visualizing data using t-SNE. J Mach Learn Res 2008;9:2579-2605.
2. van der Maaten L. Accelerating t-SNE using tree-based algorithms. J Mach Learn Res 2014;15:3221-3245.
3. Linderman GC, Rachh M, Hoskins JG, Steinerberger S, Kluger Y. Fast interpolation-based t-SNE for improved visualization of single-cell RNA-seq data. Nat Methods 2019;16:243-245.
4. Policar PG, Strazar M, Zupan B. openTSNE: a modular Python library for t-SNE dimensionality reduction and embedding. J Open Source Softw 2019;4:1576.
5. Chan DM, Rao R, Huang F, Canny JF. t-SNE-CUDA: GPU-accelerated t-SNE and its applications to modern data. arXiv 2018;1807.11824.
6. Belkina AC, Ciccolella CO, Anno R, Halpert R, Spidlen J, Snyder-Cappione JE. Automated optimized parameters for T-distributed stochastic neighbor embedding improve visualization and analysis of large datasets. Nat Commun 2019;10:5415.
7. McInnes L, Healy J, Melville J. UMAP: Uniform Manifold Approximation and Projection for Dimension Reduction. arXiv 2018;1802.03426.
8. Johnson J, Douze M, Jegou H. Billion-scale similarity search with GPUs. IEEE Trans Big Data 2021;7:535-547.

Software acknowledgements include `uwot`, `Rtsne`, FAISS, RAPIDS cuVS,
AppleSiliconFFT, `mlx-vis`, and `annembed`. See `inst/NOTICE`,
`inst/ALGORITHMIC_REFERENCES.md`, and `docs/references.md` for provenance and
license notes.
