---
title: "Getting started with fastEmbedR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fastEmbedR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`fastEmbedR` is focused on KNN-first embeddings: UMAP from precomputed nearest
neighbours, plus a native openTSNE-style optimizer. Nearest-neighbour search,
graph building, kNN models, and k-means live in the companion `faissR` package
and are re-exported as thin `fastEmbedR` wrappers.

The legacy `tsne()` and `infotsne()` entry points were removed from the public
package API. The kept path is:

1. compute or reuse KNN with `fastEmbedR::nn()`;
2. embed with `umap_knn()` / `umap()` or, explicitly, `opentsne_knn()` /
   `opentsne()`;
3. optionally place query/non-landmark points with `transform_tsne()`.

CPU-capable public functions expose `n_threads`, for example
`n_threads = 2` for a fixed two-core run. Native Apple GPU paths use
`backend = "metal"`; CUDA paths use `backend = "cuda"`. GPU requests are not
silently relabelled when unavailable.

For one-call `umap()` and `opentsne()` runs, `backend` is deliberately limited
to `"cpu"`, `"metal"`, or `"cuda"`. Matrix-input KNN is fixed internally: CPU
and Metal use FAISS CPU IVF-Flat through `faissR`, and CUDA uses FAISS GPU
IVF-Flat. To benchmark another KNN algorithm, compute it with `nn()` and pass
the result to `umap_knn()` or `opentsne_knn()`.

The package deliberately keeps the backend surface small. The local MPSGraph
FFT experiment did not improve the Metal openTSNE benchmark, so it is not
exposed as an option. The Metal path uses the package-native Objective-C++/Metal
FFT-grid implementation when compiled.

## Direct KNN Input

```{r}
library(fastEmbedR)

x <- scale(as.matrix(iris[, 1:4]))
labels <- iris$Species

knn <- fastEmbedR::nn(x, k = 31, n_threads = 2)
layout <- umap_knn(knn, n_threads = 2, seed = 4)

plot(layout, pch = 21, bg = labels)
```

`embed_knn()` dispatches to UMAP by default. Request openTSNE explicitly:

```{r}
layout2 <- embed_knn(
  knn,
  method = "opentsne",
  perplexity = 10,
  seed = 4
)
```

When iteration counts and learning rate are omitted, openTSNE uses
`auto_config = TRUE`. The native C++ policy follows opt-SNE's practical rules:
learning rate is `n / early_exaggeration`, CPU/small exact runs can end early by
monitoring KLD, and large FFT/GPU runs do not hide an expensive CPU KLD monitor
inside a GPU-labelled benchmark.

## One-Call Workflow

```{r}
fit <- umap(
  x,
  n_neighbors = 30,
  n_threads = 2,
  seed = 4
)

fit
plot(fit)

evaluate_embedding(
  x,
  fit$layout,
  labels = labels,
  k = 15,
  sample_size_for_global_metrics = nrow(x),
  sample_size_for_local_metrics = nrow(x),
  n_threads = 2
)
```

When `keep_knn = TRUE`, the non-self KNN graph used by the embedding is stored
in the returned object. The default is `FALSE` to keep large results compact.

## Native Metal UMAP

On Apple Silicon, `umap_knn(..., backend = "metal")` calls the package's
Objective-C++/Metal UMAP optimizer directly. It does not use Python,
`reticulate`, Torch, or MLX, and it does not fall back to CPU while reporting a
GPU backend.

Metal UMAP has a single optimizer path: `atomic_inplace`. This is the
visually validated native Metal edge-update kernel used for reference
comparisons; the slower or distorted experimental Metal UMAP optimizers were
removed to keep results reproducible and the user API simple.

Native openTSNE Metal is conditional: it is available only when the package is
compiled with the `knn_tsne_opentsne_metal_cpp` symbol. In that build,
`negative_gradient_method = "auto"` resolves to the package-native Metal
FFT-grid path. If the symbol is absent, `opentsne_knn(..., backend = "metal")`
fails clearly instead of running on CPU and reporting Metal.

For CPU runs, `negative_gradient_method = "auto"` resolves to
`negative_gradient_method = "fft"`, a native FIt-SNE-style grid approximation
for the negative gradient. This is useful for large two-dimensional embeddings
from a supplied KNN graph. The older Barnes-Hut route has been removed.
Native Metal FFT openTSNE uses Objective-C++/Metal kernels when compiled.
Native CUDA FFT openTSNE uses package CUDA kernels and cuFFT when compiled.

```{r, eval = FALSE}
metal_layout <- umap_knn(
  knn,
  backend = "metal",
  seed = 4
)

attr(metal_layout, "fastEmbedR_config")[c("backend", "optimizer", "graph_storage")]
```

## Landmark Workflow

`landmark_tsne()` embeds a subset with `opentsne()` and places the remaining
rows with the openTSNE-style fixed-reference transform.

When native openTSNE Metal is compiled, `backend = "metal"` can use fused Metal
projection kernels for query-to-landmark KNN, interpolation, and confidence
scoring. In this restored build, request Metal landmark t-SNE only after
checking `fastEmbedR:::metal_opentsne_native_available()`. CPU runs keep the
exact multi-threaded projection KNN for small and medium data, and switch
internally to a projection-specific candidate search only for larger
projections.

```{r, eval = FALSE}
landmark_fit <- landmark_tsne(
  x,
  landmarks = 0.5,
  n_neighbors = 30,
  perplexity = 10,
  early_exaggeration_iter = 100,
  n_iter = 250,
  transform_iter = 100,
  seed = 4
)

plot(landmark_fit)
```

UMAP supports the same landmark pattern:

```{r, eval = FALSE}
landmark_umap_fit <- landmark_umap(
  x,
  landmarks = 0.5,
  n_neighbors = 30,
  seed = 4
)

plot(landmark_umap_fit)
```

## Transform New Points

```{r, eval = FALSE}
reference <- x[1:100, , drop = FALSE]
query <- x[101:110, , drop = FALSE]

ref_fit <- opentsne(
  reference,
  perplexity = 10,
  early_exaggeration_iter = 100,
  n_iter = 250,
  standardize = FALSE,
  seed = 5
)

query_knn <- fastEmbedR::nn(reference, query, k = 15)
query_layout <- transform_tsne(
  ref_fit$layout,
  knn = query_knn,
  perplexity = 5,
  n_iter = 100,
  seed = 5
)

plot(ref_fit$layout, pch = 21, bg = labels[1:100])
points(query_layout, pch = 24, bg = labels[101:110], cex = 1.2)
```

## Backend Reporting

`backend_info()` reports optional CUDA, Metal, FAISS, and cuVS availability.
Requested unavailable GPU/accelerator backends fail clearly rather than being
silently replaced by CPU and reported as GPU.

```{r}
backend_info()
```

## Reproducing the MNIST Benchmark

The benchmark helper can use raw flattened MNIST images. This avoids measuring
a PCA-preprocessed problem when the goal is to test the full image matrix:

```{sh, eval = FALSE}
Rscript tools/benchmark_mnist70k_current_backends.R \
  --feature-source=raw \
  --n=70000 \
  --k=50 \
  --seed=6 \
  --threads=4 \
  --backends=cpu,metal \
  --run-uwot=true \
  --run-umap=true \
  --run-opentsne=true \
  --run-rtsne=true \
  --run-landmark=true
```

The output table separates nearest-neighbour time, embedding time, and
landmark projection/transform time where the implementations expose those
stages. Reference packages that do not expose stage timing are reported with
their total call time in the closest matching column and a note in the saved
CSV.
