---
title: "fastEmbedR: native UMAP and t-SNE workflows"
author: "Stefano Cacciatore"
output:
    rmarkdown::html_vignette:
        toc: true
        number_sections: true
vignette: >
    %\VignetteIndexEntry{fastEmbedR: native UMAP and t-SNE workflows}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.align = "center",
    fig.width = 6.2,
    fig.height = 3.2,
    out.width = "92%",
    fig.retina = if (knitr::is_html_output()) 2 else NULL,
    dpi = 96,
    dev.args = list(bg = "white")
)
options(width = 90)
set.seed(4)

embedding_colors <- c("#2678C8", "#E4576B", "#26A269")

plot_embedding <- function(layout, labels, main) {
    labels <- factor(labels)
    plot(
        as.matrix(layout),
        col = embedding_colors[as.integer(labels)],
        pch = 19,
        cex = 0.75,
        asp = 1,
        axes = FALSE,
        xlab = "",
        ylab = "",
        main = main
    )
    box(col = "#B8B8B8")
}
```

# Overview

`fastEmbedR` provides native implementations of two widely used nonlinear
dimension-reduction workflows:

- UMAP from a sparse nearest-neighbor graph;
- interpolation-based t-SNE with sparse attractive forces and an FFT-grid
    approximation for repulsion.

The same R interface selects native CPU, Apple Metal, or NVIDIA CUDA code.
One-call functions include nearest-neighbor search, initialization, graph or
affinity construction, and optimization. KNN-input functions let an existing
neighbor graph be reused without repeating search. Explicit GPU requests either
run on the requested GPU backend or fail with an informative error; they are
never silently reported as GPU work after falling back to CPU.

## Select the backend and CPU core count

Every backend-capable function accepts `backend = "cpu"`, `"metal"`, or
`"cuda"`. An explicit function argument is the clearest choice for a
reproducible script:

```r
fit_cpu <- fastEmbedR::umap(
    x,
    backend = "cpu",
    n.cores = 4,
    seed = 4
)

tsne_cpu <- fastEmbedR::tsne(
    x,
    perplexity = 10,
    backend = "cpu",
    n.cores = 4,
    seed = 4
)

fit_metal <- fastEmbedR::umap(x, backend = "metal", seed = 4)
fit_cuda <- fastEmbedR::umap(x, backend = "cuda", seed = 4)
```

`n.cores` is a positive integer controlling CPU parallel work, including the
native nearest-neighbor search and the CPU graph, affinity, and optimization
stages that support parallel execution. For UMAP, the optimizer currently uses
at most four workers and records the requested and effective values. When
`n.cores` is omitted, KNN uses one worker, UMAP selects a size-aware default
from one to four workers, and t-SNE uses its thread option, which defaults to
four workers. Metal and CUDA execute on their respective accelerators and
ignore `n.cores`; the argument does not control GPU concurrency.

For an interactive session, set a default once and omit `backend` from later
calls:

```r
old_backend <- getOption("fastEmbedR.backend")

fastEmbedR::fastEmbedR_backend("metal")
fastEmbedR::fastEmbedR_backend()       # reports "metal"
u <- fastEmbedR::umap(x)               # uses Metal
t <- fastEmbedR::tsne(x, perplexity = 10)

u.cpu <- fastEmbedR::umap(             # explicit arguments take precedence
    x,
    backend = "cpu",
    n.cores = 4
)

options(fastEmbedR.backend = old_backend)
```

The equivalent base-R setting is
`options(fastEmbedR.backend = "cpu")`, `"metal"`, or `"cuda"`. For a batch
job, set `FASTEMBEDR_BACKEND` before starting R, for example
`FASTEMBEDR_BACKEND=cuda Rscript analysis.R`. Resolution precedence is:

1. the explicit `backend` function argument;
2. `options(fastEmbedR.backend = ...)` or `fastEmbedR_backend()`;
3. the `FASTEMBEDR_BACKEND` environment variable;
4. CPU when none of the above is set.

Before requesting an accelerator, inspect the installed build and runtime
device support:

```r
fastEmbedR::fastEmbedR_capabilities()[
    , c("backend", "available", "device", "knn_engine", "unavailable_reason")
]
```

The backend requested by the user and the backend that actually ran are also
recorded in returned objects, for example
`fit_cpu$parameters[c("backend", "nn_backend", "n.cores")]`. A requested
Metal or CUDA backend must be available and execute natively; otherwise the
function stops with an informative error instead of silently using CPU.

The public workflow is intentionally small:

| Goal | Function |
| --- | --- |
| Precompute native non-self neighbors | `precompute_knn()` |
| Run UMAP from a data matrix | `umap()` |
| Run interpolation-based t-SNE from a data matrix | `tsne()` |
| Reuse a KNN graph | `umap_knn()` or `tsne_knn()` |
| Compute reusable PCA scores or t-SNE initialization | `pca()` |
| Approximate a very large run with landmarks | Landmark functions |
| Place new observations in a fixed t-SNE reference | `transform_tsne()` |
| Quantify embedding quality | `evaluate_embedding()` |

`fastEmbedR_api()` is the normative inventory for the installed release. It
separates these canonical workflows from advanced prepared-state interfaces,
diagnostics, secondary graph utilities, and the compatibility dispatcher, and
reports accepted/returned classes, backend support, residency, related methods,
and lifecycle status.

```{r api-map}
api <- fastEmbedR::fastEmbedR_api()
api[api$tier == "canonical", c("function", "purpose", "return_class")]
```

Because other R packages export generic names such as `umap()` and `pca()`, use
`fastEmbedR::umap()` and `fastEmbedR::pca()` in scripts, package code, and
published analyses. fastEmbedR follows ordinary R search-path rules and does
not silently resolve masking conflicts.

# A first analysis

We use the four numeric Iris measurements for a small, reproducible example.
Iris is supplied by R's separate `datasets` package and is not bundled with
fastEmbedR. The matrix is scaled once outside the embedding calls, so
`standardize = FALSE` prevents repeated preprocessing.

```{r data}
library(fastEmbedR)

iris_data <- datasets::iris
x <- scale(as.matrix(iris_data[, 1:4]))
labels <- iris_data$Species
dim(x)
```

UMAP uses a neighborhood size, whereas t-SNE uses perplexity. For this small
dataset we use 15 neighbors and perplexity 10. Missing t-SNE optimization
settings are chosen by the package's native auto-configuration policy. Its
default initialization is a small-scale randomized-PCA layout.

```{r first-embeddings}
fit_umap <- umap(
    x,
    n_neighbors = 15,
    graph_mode = "fuzzy",
    standardize = FALSE,
    n.cores = 2,
    seed = 4
)

fit_tsne <- tsne(
    x,
    perplexity = 10,
    standardize = FALSE,
    n.cores = 2,
    seed = 4
)

c(
    UMAP_seconds = fit_umap$metrics$elapsed,
    tsne_seconds = fit_tsne$metrics$elapsed
)
```

Both functions return a `fastEmbedR_embedding` object. The full coordinates are
in `$layout`; `$metrics`, `$timings`, and `$parameters` preserve enough context
to audit a run.

```{r first-embedding-plots, fig.cap="Iris one-call CPU embeddings."}
old_par <- par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1), bg = "white")
plot_embedding(fit_tsne$layout, labels, "t-SNE")
plot_embedding(fit_umap$layout, labels, "UMAP (fuzzy graph)")
par(old_par)
```

The backend recorded in the result is the backend that actually ran:

```{r result-metadata}
fit_umap$parameters[c("backend", "nn_backend", "n.cores", "graph_mode")]
fit_tsne$parameters[c(
    "backend", "nn_backend", "n.cores", "perplexity",
    "affinity_support", "affinity_support_k"
)]
```

# Parameter philosophy and intended scope

The two embedding functions deliberately expose different levels of control.
t-SNE exposes perplexity/support, initialization, iteration counts, early
and normal exaggeration, learning rate, momentum, clipping, and exact-versus-FFT
repulsion. Set `auto_config = FALSE` and provide explicit values when automatic
iteration or stopping choices are not appropriate.

UMAP is an opinionated high-throughput implementation, not a general-purpose
hyperparameter interface. Users select
`n_neighbors`, metric, graph mode, preprocessing, backend, seed, and CPU thread
count. The release policy selects the remaining values and records them in
`fit$parameters`: 500 epochs below 10,000 observations and 200 for larger
ordinary profiles (at least 300 for high-variability profiles), five negative
samples, spread 1, repulsion strength 1, `min_dist` 0.01 or 0.1 under the
wide-shell rule, and learning rate 1 or 1.25 under the same rule.

`umap_init()` makes package-native initialization reusable, but `umap()` does
not accept arbitrary initial coordinates. The nearest-neighbor router targets
recall 0.99 and intentionally hides index tuning; supply an external KNN object
when search policy must be controlled independently. Fixed seeds do not ensure
bitwise Metal or CUDA identity because atomic update order can differ.

These settings affect cluster compactness, convergence effort, and the balance
between attractive and repulsive updates. Fixing them keeps one tested policy
aligned across CPU, Metal, and CUDA; it does not imply that the selected values
are universally optimal. Analyses requiring `min_dist`, spread, epoch,
learning-rate, repulsion, negative-sample, or optimizer-mode sweeps should use
a general-purpose implementation such as `uwot` or `umap-learn`. Float32
input, reusable KNN, prepared graphs, compact t-SNE support, and
landmarking remain explicit memory/speed controls; compact support and
landmarking must be described as approximations.

# Choosing between UMAP and t-SNE

The two methods answer related but different questions.

| Property | UMAP | interpolation-based t-SNE |
| --- | --- | --- |
| Model | Weighted or binary KNN graph | Sparse affinities |
| Main scale parameter | `n_neighbors` | `perplexity` |
| Initialization | Graph-native initialization | Randomized PCA by default |
| Repulsion | Sampled negative edges | Exact or FFT-grid negative gradient |
| New data | Landmark projection | Fixed-reference transform |

UMAP emphasizes local neighborhoods with some graph-level organization.
t-SNE emphasizes local probability neighborhoods and cluster organization.

Distances and empty space in either two-dimensional plot should not be treated
as calibrated biological effect sizes. Compare methods with neighborhood and
label-aware metrics rather than visual appearance alone.

# Optional downstream graph clustering

Graph clustering is a secondary downstream utility; the primary package
contribution is native UMAP and t-SNE embedding.

`knn_graph()` builds a compact undirected graph from raw data, an embedding, or
precomputed neighbors. It uses the existing KNN route selected by `backend`;
there is no second nearest-neighbor method menu inside the graph API.

```{r graph-clustering}
graph <- knn_graph(fit_tsne, k = 20, weight = "snn", n.cores = 2)
communities <- graph_cluster(graph, method = "leiden", seed = 4)
table(communities$membership)
```

Louvain and Leiden have native CPU, CUDA, and Metal backends and use sparse
graph storage. Select an accelerator explicitly with
`graph_cluster(..., backend = "cuda")` or `"metal"`; an unavailable backend
fails rather than falling back to CPU. `walktrap` is the true Pons-Latapy
random-walk agglomeration method and remains CPU-only because its exact
transition matrix is intentionally limited to moderate graphs. Neither
`igraph`, cuGraph, nor an external clustering routine is called at run time.

# PCA and reproducible t-SNE initialization

`pca()` exposes the backend-native truncated PCA used for initialization. It
deliberately omits an SVD-method menu: randomized SVD is the supported
decomposition family. For `backend = "cpu"`,
`n.cores` sets the CPU BLAS/OpenMP thread limit for the call.
`tsne_init = FALSE` is the default; request it when the t-SNE-ready layout
is needed. `irlba` is not used by the package implementation and appears only
as an external benchmark reference.

```{r pca-initialization}
pca_fit <- pca(
    x,
    ncomp = 2,
    center = TRUE,
    scale = FALSE,
    backend = "cpu",
    n.cores = 2,
    seed = 4,
    tsne_init = TRUE
)

c(
    score_rows = nrow(pca_fit$scores),
    score_columns = ncol(pca_fit$scores),
    init_max_sd = max(apply(as.matrix(pca_fit$tsne_init), 2, stats::sd))
)
```

The ordinary scores remain in `pca_fit$scores`. The added
`pca_fit$tsne_init` is centered and rescaled so the largest component
standard deviation is `1e-4`, the small initialization scale expected by
t-SNE optimizers. CPU fits record the requested and observable thread counts
in `n.cores_requested` and `n.cores_effective`. A single-threaded BLAS may
still report one effective thread. The initialization can be reused across
methods, seeds, or backends:

```{r reuse-initialization}
layout_tsne_fixed_init <- tsne(
    x,
    perplexity = 10,
    Y_init = pca_fit$tsne_init,
    standardize = FALSE,
    n.cores = 2,
    seed = 4
)
```

CPU PCA uses the package-native float32 blocked RSVD path. Metal uses a native
float32 block-subspace TSVD with Metal Performance Shaders matrix products.
CUDA uses native RAPIDS RAFT TSVD when that support is compiled. No backend
calls Python or `reticulate`. Float32 CUDA input is passed to RAFT without an
intermediate R double matrix, and float32 scores and loadings are preserved.

# Reusing a nearest-neighbor graph

The KNN-input API is useful when comparing embeddings, graph modes, or seeds.
`precompute_knn()` exposes the same backend-specific native search selected by
the one-call embedding functions, without exposing a second algorithm menu.

```{r precompute-knn}
knn <- precompute_knn(
    x,
    k = 30,
    metric = "euclidean",
    backend = "cpu",
    n.cores = 2
)
knn
```

`prepare_umap_knn()` builds graph state and `epochs_per_sample` once.
`prepare_tsne_knn()` validates, strips self-neighbors when needed, and
stores the compact affinity support. It uses one candidate neighbor per unit
of perplexity and trims wider KNN inputs. Prepared objects are useful for
repeated runs.

```{r prepared-knn}
prepared_umap <- prepare_umap_knn(
    knn,
    graph_mode = "fuzzy",
    n.cores = 2
)
prepared_tsne <- prepare_tsne_knn(knn, perplexity = 10)

layout_umap_reused <- umap_knn(
    prepared_umap,
    n.cores = 2,
    seed = 4
)
layout_tsne_reused <- tsne_knn(
    prepared_tsne,
    Y_init = pca_fit$tsne_init,
    n.cores = 2,
    seed = 4
)

cbind(UMAP = dim(layout_umap_reused), TSNE = dim(layout_tsne_reused))
```

## Independent UMAP initialization

`umap_init()` separates graph construction and sparse initialization from
layout optimization. The returned object retains the exact prepared graph and
initial coordinates, so `umap_knn()` can reuse both without rebuilding either
stage. This is useful for repeated seeds, backend agreement tests, and
optimizer-only timing.

```{r independent-umap-init}
umap_initialization <- umap_init(
    prepared_umap,
    backend = "cpu",
    graph_mode = "fuzzy",
    n.cores = 2,
    seed = 4
)

layout_from_initialization <- umap_knn(
    umap_initialization,
    backend = "cpu",
    n.cores = 2,
    seed = 4
)

umap_initialization$timings
```

The initialization object can also be passed to a Metal or CUDA optimizer.
That controlled comparison holds the graph and starting coordinates fixed;
ordinary one-call GPU embeddings continue to use their resident backend path.

# UMAP graph modes

The graph mode is an explicit scientific choice, not a hidden speed setting.

- `graph_mode = "fuzzy"` is the default and uses locally scaled fuzzy
    membership strengths from the standard UMAP graph construction.
- `graph_mode = "binary"` symmetrizes unit-weight neighbor edges and then uses
    the same UMAP optimizer. It is an explicit sensitivity mode and can produce
    stronger separation on some datasets.

```{r graph-modes}
layout_binary <- umap_knn(
    knn,
    graph_mode = "binary",
    n.cores = 2,
    seed = 4
)
layout_fuzzy <- umap_knn(
    knn,
    graph_mode = "fuzzy",
    n.cores = 2,
    seed = 4
)
```

```{r graph-mode-plots, fig.cap="UMAP graph modes use the same optimizer."}
old_par <- par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1), bg = "white")
plot_embedding(layout_binary, labels, "Binary graph")
plot_embedding(layout_fuzzy, labels, "Fuzzy graph")
par(old_par)
```

Record the chosen mode whenever embeddings are reported or compared.

# Float32 and memory-conscious analysis

All native embedding buffers use float32 and integer indices. A standard R
numeric matrix is converted at the native boundary. Supplying a
`float::float32` matrix avoids retaining an unnecessary double-precision copy
in R, and the returned layout remains float32.

```{r float32, results='hold'}
if (requireNamespace("float", quietly = TRUE)) {
    x_float <- float::fl(x)
    fit_float <- umap(
        x_float,
        n_neighbors = 15,
        graph_mode = "fuzzy",
        standardize = FALSE,
        n.cores = 2,
        seed = 4
    )
    c(
        input_class = class(x_float)[1],
        layout_class = class(fit_float$layout)[1]
    )
}
```

For large datasets:

1. perform scaling or feature extraction once and reuse the result;
2. use float32 input when practical;
3. leave `keep_knn = FALSE` unless neighbor matrices are needed later;
4. use prepared KNN/graph objects for repeated seeds;
5. consider landmarking only when its approximation is scientifically
    acceptable.

# Embedding new observations

`transform_tsne()` places query observations into an existing reference
embedding without moving the reference points. It builds query-to-reference
affinities, initializes each query near its reference neighbors, and optimizes
only the query coordinates. Reference and query data must have undergone the
same preprocessing.

For this executable example, every fifth Iris observation is treated as new.
We compute its exact query-to-reference KNN directly. The same API accepts any
plain host list with query-to-reference `indices` and `distances`.

```{r transform-new-data}
query_rows <- seq(5L, nrow(x), by = 5L)
reference_rows <- setdiff(seq_len(nrow(x)), query_rows)
reference_data <- x[reference_rows, , drop = FALSE]
query_data <- x[query_rows, , drop = FALSE]

reference_fit <- tsne(
    reference_data,
    perplexity = 10,
    standardize = FALSE,
    n.cores = 2,
    seed = 4
)

exact_query_knn <- function(reference, query, k) {
    squared_distances <- outer(
        rowSums(query^2), rowSums(reference^2), "+"
    ) - 2 * tcrossprod(query, reference)
    squared_distances[squared_distances < 0] <- 0
    indices <- t(apply(squared_distances, 1L, order))[
        , seq_len(k),
        drop = FALSE
    ]
    distances <- matrix(
        sqrt(squared_distances[
            cbind(rep(seq_len(nrow(query)), each = k), as.vector(t(indices)))
        ]),
        nrow = nrow(query),
        byrow = TRUE
    )
    list(indices = indices, distances = distances)
}

query_knn <- exact_query_knn(reference_data, query_data, k = 15)
query_layout <- transform_tsne(
    reference_fit,
    knn = query_knn,
    perplexity = 5,
    n_iter = 100,
    n.cores = 2,
    backend = "cpu",
    seed = 4
)
```

```{r transform-plot, fig.cap="Transform into a fixed t-SNE reference."}
reference_labels <- factor(labels[reference_rows], levels = levels(labels))
query_labels <- factor(labels[query_rows], levels = levels(labels))
par(bg = "white")
plot(
    reference_fit$layout,
    col = embedding_colors[as.integer(reference_labels)],
    pch = 19,
    cex = 0.65,
    asp = 1,
    axes = FALSE,
    xlab = "",
    ylab = "",
    main = "Fixed-reference transform"
)
points(
    query_layout,
    col = embedding_colors[as.integer(query_labels)],
    pch = 17,
    cex = 1.05
)
legend(
    "bottomright",
    legend = c("Reference", "New"),
    pch = c(19, 17),
    bty = "n",
    cex = 0.8
)
box(col = "#B8B8B8")
```

For Metal or CUDA, request the same backend explicitly. If the native
fixed-reference path is unavailable, the call fails rather than relabeling a
CPU transform as GPU work.

# Landmarking

Landmarking embeds a representative subset and places the remaining rows in
that reference. It is never activated silently by `umap()` or `tsne()`.
The final layout still has one row per input observation. Reference and
projection searches use fastEmbedR's native KNN routes.

The staged API keeps selection, reference fitting, and projection separate.
This is useful when the same landmark reference will be reused:

```{r staged-landmarking}
landmark_selection <- select_landmarks(
    x,
    landmarks = 0.5,
    seed = 4,
    n.cores = 2
)

umap_reference <- fit_landmark_model(
    x,
    landmark_selection,
    method = "umap",
    n_neighbors = 15,
    graph_mode = "fuzzy",
    backend = "cpu",
    n.cores = 2,
    seed = 4
)

umap_projected <- project_landmark_model(
    umap_reference,
    x,
    refinement_epochs = 50,
    n.cores = 2
)
```

`fit_landmark_model()` invokes the ordinary production `umap()` or
`tsne()` implementation on the reference rows. Consequently, UMAP
`graph_mode`, optimizer settings, seed, and backend are explicit and
consistent with a full-data fit. `project_landmark_model()` computes only the
query-to-reference neighbors. It does not recompute neighbors among
landmarks or among query observations. Passing a matrix of new observations
instead of `x` projects those observations into the same fixed reference
embedding.

The one-call functions below perform the same workflow when the intermediate
selection and model do not need to be retained.

```{r landmarking}
fit_landmark_umap <- landmark_umap(
    x,
    landmarks = 0.5,
    n_neighbors = 15,
    graph_mode = "fuzzy",
    standardize = FALSE,
    n.cores = 2,
    seed = 4
)

fit_landmark_tsne <- landmark_tsne(
    x,
    landmarks = 0.5,
    perplexity = 10,
    standardize = FALSE,
    n.cores = 2,
    seed = 4
)

rbind(
    UMAP = fit_landmark_umap$metrics[
        1, c(
            "elapsed", "reference_embedding_elapsed",
            "landmark_projection_knn_elapsed", "landmark_refinement_elapsed"
        )
    ],
    TSNE = setNames(
        fit_landmark_tsne$metrics[
            1, c(
                "elapsed", "reference_embedding_elapsed",
                "landmark_projection_knn_elapsed", "transform_elapsed"
            )
        ],
        c(
            "elapsed", "reference_embedding_elapsed",
            "landmark_projection_knn_elapsed", "landmark_refinement_elapsed"
        )
    )
)
```

```{r landmark-plots, fig.cap="Iris with 50% landmarks."}
old_par <- par(mfrow = c(1, 2), mar = c(1, 1, 2.5, 1), bg = "white")
plot_embedding(fit_landmark_tsne$layout, labels, "Landmark t-SNE")
plot_embedding(fit_landmark_umap$layout, labels, "Landmark UMAP")
par(old_par)
```

CPU uses native HNSW for query-to-reference search. Metal uses native exact
search for small references and recall-tuned IVF-Flat for larger references,
followed by fixed-reference transforms. CUDA uses native exact or IVF-Flat
query search and keeps KNN, layouts, and refinement buffers resident on the
GPU. The result records landmark count, fraction, projection backend, and
separate reference, query-KNN, and transform timings so a landmark
approximation cannot be mistaken for a full embedding.

# Measuring embedding quality

`evaluate_embedding()` reports local, global, and label-aware metrics. For
large datasets it uses deterministic subsamples rather than constructing a
complete all-pairs distance matrix. Here all Iris observations are small enough
to evaluate directly. Supplying `reference_nn` also avoids repeating the
high-dimensional neighbor search.

```{r quality}
quality_umap <- evaluate_embedding(
    x,
    fit_umap$layout,
    labels = labels,
    k = 10,
    primary_k = 10,
    reference_nn = knn,
    sample_size_for_local_metrics = nrow(x),
    sample_size_for_global_metrics = nrow(x),
    seed = 4,
    method = "UMAP",
    backend = "cpu",
    dataset = "iris",
    n.cores = 2
)

quality_tsne <- evaluate_embedding(
    x,
    fit_tsne$layout,
    labels = labels,
    k = 10,
    primary_k = 10,
    reference_nn = knn,
    sample_size_for_local_metrics = nrow(x),
    sample_size_for_global_metrics = nrow(x),
    seed = 4,
    method = "t-SNE",
    backend = "cpu",
    dataset = "iris",
    n.cores = 2
)

quality <- rbind(quality_umap, quality_tsne)
quality[, c(
    "method", "trustworthiness", "knn_preservation_10",
    "distance_spearman", "silhouette", "label_knn_accuracy"
)]
```

No single score is sufficient. Trustworthiness and neighbor preservation
measure local structure; distance correlation is a coarse global diagnostic;
silhouette and KNN label accuracy are meaningful only when labels are valid
external annotations rather than tuning targets.

# CPU, Metal, and CUDA

The backend argument has exactly three choices.

| Backend | One-call KNN | Embedding implementation |
| --- | --- | --- |
| `"cpu"` | HNSW | Multithreaded C++ |
| `"metal"` | Exact or IVF-Flat | Metal kernels |
| `"cuda"` | Exact or IVF-Flat | CUDA kernels |

CPU t-SNE uses a fast Fourier transform (FFT) grid. Metal uses native
Objective-C++ kernels. CUDA uses FAISS GPU or RAPIDS cuVS search and keeps
embedding buffers resident on the device.

GPU use is explicit:

```{r gpu-examples, eval=FALSE}
fit_metal <- tsne(x, perplexity = 10, backend = "metal", seed = 4)
fit_cuda <- umap(x, n_neighbors = 15, backend = "cuda", seed = 4)
```

If the requested backend was not compiled or is unavailable at runtime, these
calls stop. CPU fallback is not relabeled as Metal or CUDA. Because parallel
reductions and atomic updates can be ordered differently, fixed seeds provide
controlled stochastic initialization and sampling but do not promise bitwise
identity between devices or thread counts.

## Package boundary

The package owns the matrix and KNN-input interfaces, native nearest-neighbor
routes, PCA initialization, graph or affinity construction, low-dimensional
optimization, landmark transformation, graph construction, community
detection, quality metrics, and returned metadata. A plain host KNN object can
be supplied as `indices` and `distances`, while a CUDA KNN result produced by
`precompute_knn()` can remain device-resident through the CUDA embedding path.
No external R nearest-neighbor package is called by the production workflow.

## UMAP implementation

The fuzzy path follows the graph construction and cross-entropy optimization
defined by McInnes, Healy, and Melville. `fastEmbedR` changes the execution
boundary rather than the UMAP objective: graph weights, coordinates, schedules,
and optimizer state use compact float32 buffers, and dense graph intermediates
are avoided. CPU parallelizes row preparation and sampled-edge updates; Metal
retains graph and optimizer buffers in unified memory; CUDA constructs and
optimizes the graph from device-resident KNN arrays.

`graph_mode = "fuzzy"` is the standard default. The explicit `"binary"` mode
assigns unit weight to the symmetric KNN union and is an adjacency-only
sensitivity analysis for settings where neighbor identity is more reliable than
distance calibration. It is not standard UMAP and is not guaranteed to be
faster because every retained edge receives the maximum sampling frequency.

## Interpolation-based t-SNE implementation

The affinity definition and Kullback--Leibler objective follow van der Maaten
and Hinton. Interpolation-based repulsion follows FIt-SNE, while the automatic
two-phase optimization policy is informed by opt-SNE and related t-SNE
implementations. The compact policy uses `ceiling(perplexity)` non-self
affinity candidates.

CPU uses a retained FFT-grid workspace, Metal uses package-native scatter,
transform, gather, attraction, and optimizer kernels, and CUDA combines native
kernels with cuFFT and persistent device buffers. PCA initialization is created
by `pca(..., tsne_init = TRUE)` or internally by the one-call workflow. The
fixed-reference transform follows the published reference-space embedding
procedure and keeps the fitted reference coordinates unchanged.

## Graph clustering and landmarking

`knn_graph()` builds weighted sparse graphs using shared-neighbor, inverse-
distance, or binary edges. `graph_cluster()` provides native Louvain, Leiden,
and exact CPU Walktrap implementations. Louvain and Leiden have CPU, Metal, and
CUDA routes; Walktrap remains CPU-only because its exact transition
representation is quadratic in the number of vertices. The implementations are
tested against `igraph` on matched weighted graphs, but `igraph` is not a
runtime dependency.

Landmark workflows separate selection, reference embedding, projection KNN,
local correction, and optional fixed-reference refinement. They are explicit
approximations: timings and returned parameters identify landmark and projection
stages rather than presenting them as ordinary full-data embeddings.

## Backend reporting and software acknowledgements

An explicit accelerator request must execute on that backend or fail. Returned
objects record the resolved embedding backend, nearest-neighbor route,
precision,
parameters, and stage timings. Backend availability and diagnostic reasons are
reported by `fastEmbedR_capabilities()`.

The package is distributed under the MIT license. Adapted source and linked
library boundaries, upstream commits, SPDX identifiers, and retained notices
are recorded in `inst/NOTICE`, `inst/COPYRIGHTS`,
`inst/THIRD_PARTY_DEPENDENCIES.json`, and the repository's source-provenance
documentation. Software used only for validation remains outside the production
call graph.

# A practical analysis checklist

Before publishing an embedding analysis:

1. state whether input features were centered, scaled, or reduced by PCA;
2. report method, backend, seed, `n_neighbors` or perplexity, and UMAP graph
    mode;
3. distinguish full runs from landmark approximations;
4. compare total runtime when competing packages expose different internal
    boundaries;
5. report trustworthiness and neighbor preservation alongside any label-aware
    score;
6. inspect the plot for outliers, splitting, crowding, and seed sensitivity;
7. save `sessionInfo()` and the result's `$parameters` and `$timings`.

# References

1. van der Maaten L, Hinton G. Visualizing data using t-SNE. J Mach Learn
    Res 2008;9:2579-2605.
2. van der Maaten L. Accelerating t-SNE using tree-based algorithms. J Mach
    Learn Res 2014;15:3221-3245.
3. Linderman GC, Rachh M, Hoskins JG, Steinerberger S, Kluger Y. Fast
    interpolation-based t-SNE for improved visualization of single-cell
    RNA-seq data. Nat Methods 2019;16:243-245.
4. Policar PG, Strazar M, Zupan B. openTSNE: a modular Python library for
    t-SNE dimensionality reduction and embedding. J Stat Softw 2024;109:1-30.
5. Chan DM, Rao R, Huang F, Canny JF. t-SNE-CUDA: GPU-accelerated t-SNE and
    its applications to modern data. arXiv 2018;1807.11824.
6. Belkina AC, Ciccolella CO, Anno R, Halpert R, Spidlen J,
    Snyder-Cappione JE. Automated optimized parameters for T-distributed
    stochastic neighbor embedding improve visualization and analysis of
    large datasets. Nat Commun 2019;10:5415.
7. Policar PG, Strazar M, Zupan B. Embedding to reference t-SNE space
    addresses batch effects in single-cell classification. Mach Learn
    2021;110:1211-1231.
8. McInnes L, Healy J, Melville J. UMAP: Uniform Manifold Approximation and
    Projection for Dimension Reduction. arXiv 2018;1802.03426.
9. Johnson J, Douze M, Jegou H. Billion-scale similarity search with GPUs.
    IEEE Trans Big Data 2021;7:535-547.
10. Blondel VD, Guillaume JL, Lambiotte R, Lefebvre E. Fast unfolding of
    communities in large networks. J Stat Mech 2008;2008:P10008.
11. Traag VA, Waltman L, van Eck NJ. From Louvain to Leiden: guaranteeing
    well-connected communities. Sci Rep 2019;9:5233.
12. Pons P, Latapy M. Computing communities in large networks using random
    walks. J Graph Algorithms Appl 2006;10:191-218.

```{r session-info}
sessionInfo()
```
