---
title: "Getting Started With fastEmbedR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started With fastEmbedR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`fastEmbedR` focuses on fast UMAP from nearest-neighbour matrices. The two most
important functions are `nn()` and `embed_knn()`.

```{r}
library(fastEmbedR)

x <- scale(as.matrix(iris[, 1:4]))
labels <- iris$Species

knn <- nn(x, k = 16)
layout <- embed_knn(knn, method = "umap", seed = 4)
head(layout)
```

The same workflow can be run with `umap()`, which computes KNN internally.

```{r}
fit <- umap(
  x,
  labels = labels,
  n_neighbors = 15,
  seed = 4,
  silhouette_sample = NULL,
  preserve_sample = NULL
)
fit$metrics[, c("method", "n", "n_neighbors", "elapsed")]
```

If fastEmbedR was built against the FAISS C++ library, `nn()` can use FAISS
directly:

```{r, eval = FALSE}
faiss_available()

knn_faiss <- nn(x, k = 16, backend = "faiss")
knn_faiss_ivf <- nn(x, k = 16, backend = "faiss_ivf")
```

These are real FAISS library calls. They are deliberately separate from
`backend = "cpu_faiss_ivf"`, which is the package-native FAISS-style IVF
approximation.

To compare with `uwot`, use the same KNN and the same UMAP settings:

```{r, eval = FALSE}
library(uwot)

idx <- knn$indices[, -1, drop = FALSE]
dst <- knn$distances[, -1, drop = FALSE]

fast <- fastEmbedR::embed_knn(list(indices = idx, distances = dst), seed = 4)
ref <- uwot::umap(
  x,
  nn_method = list(idx = idx, dist = dst),
  n_neighbors = ncol(idx),
  n_epochs = 200,
  min_dist = 0.01,
  negative_sample_rate = 5,
  learning_rate = 1,
  fast_sgd = TRUE,
  n_threads = 4,
  n_sgd_threads = 4
)
```

For large landmark embeddings, you can opt into confidence-gated refinement.
The package first embeds landmarks, projects all rows, then refines only rows
whose projection weights are diffuse or far from their nearest landmark.

```{r, eval = FALSE}
options(fastEmbedR.selective_landmark_refinement = TRUE)

fit <- umap(
  x_large,
  landmarks = 0.5,
  seed = 4,
  backend = "cpu"
)

fit$parameters[c(
  "landmark_refinement_policy",
  "landmark_refinement_selected",
  "landmark_refinement_backend"
)]
```

The default remains full landmark refinement because the selective mode can
reduce refinement work but should be checked for each dataset's neighbourhood
quality.

## Landmark t-SNE

`landmark_tsne()` follows the same idea used by openTSNE for adding new points:
embed a reference subset, initialize non-landmarks from nearby reference
coordinates, then optimize each non-landmark against the fixed reference map.
The implementation is native fastEmbedR code and does not call Python.

```{r}
tsne_landmark <- landmark_tsne(
  x,
  labels = labels,
  landmarks = 0.5,
  reference_method = "infotsne",
  n_neighbors = 15,
  perplexity = 5,
  max_iter = 25,
  n_negatives = 10,
  transform_iter = 10,
  transform_n_negatives = 10,
  standardize = FALSE,
  silhouette_sample = NULL,
  preserve_sample = NULL
)

tsne_landmark$metrics[, c(
  "method",
  "reference_method",
  "n",
  "n_landmarks",
  "elapsed",
  "reference_embedding_elapsed",
  "landmark_projection_knn_elapsed",
  "transform_elapsed"
)]
```

When a reference embedding is already available, `transform_tsne()` can be used
directly with a query-to-reference KNN object.

```{r}
ref <- x[1:100, , drop = FALSE]
qry <- x[101:110, , drop = FALSE]
ref_fit <- infotsne(
  ref,
  n_neighbors = 15,
  perplexity = 5,
  max_iter = 25,
  n_negatives = 10,
  standardize = FALSE,
  silhouette_sample = NULL,
  preserve_sample = NULL
)
query_knn <- nn(ref, qry, k = 15, backend = "cpu")
query_layout <- transform_tsne(
  ref_fit$layout,
  query_knn,
  perplexity = 5,
  n_iter = 10,
  n_negatives = 10
)
head(query_layout)
```

GPU backends are explicit:

```{r}
backend_info()
```

If CUDA or Metal is requested but unavailable, `fastEmbedR` does not silently
run CPU code and label it as GPU.
