---
title: "Getting started with fastEmbedR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fastEmbedR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`fastEmbedR` is focused on KNN-first embeddings: UMAP from precomputed nearest
neighbours, plus a native openTSNE-style CPU optimizer.

The legacy `tsne()` and `infotsne()` entry points were removed from the public
package API. The kept path is:

1. compute or reuse KNN with `nn()`;
2. embed with `umap_knn()` / `umap()` or, explicitly, `opentsne_knn()` /
   `opentsne()`;
3. optionally place query/non-landmark points with `transform_tsne()`.

CPU-capable public functions expose `n_threads`, for example
`n_threads = 4` for a fixed four-core run. Native Apple GPU paths use
`backend = "metal"`; `backend = "gpu"` is a convenience request for a real GPU
path and is not silently relabelled when unavailable.

## Direct KNN Input

```{r}
library(fastEmbedR)

x <- scale(as.matrix(iris[, 1:4]))
labels <- iris$Species

knn <- nn(x, k = 31)
layout <- umap_knn(knn, n_threads = 4, seed = 4)

plot(layout, pch = 21, bg = labels)
```

`embed_knn()` dispatches to UMAP by default. Request openTSNE explicitly:

```{r}
layout2 <- embed_knn(
  knn,
  method = "opentsne",
  perplexity = 10,
  early_exaggeration_iter = 100,
  n_iter = 250,
  seed = 4
)
```

## One-Call Workflow

```{r}
fit <- umap(
  x,
  labels = labels,
  n_neighbors = 30,
  n_threads = 4,
  seed = 4,
  silhouette_sample = NULL,
  preserve_sample = NULL
)

fit
plot(fit)
```

When `keep_knn = TRUE`, the non-self KNN graph used by the embedding is stored
in the returned object. The default is `FALSE` to keep large results compact.

## Native Metal UMAP

On Apple Silicon, `umap_knn(..., backend = "metal")` calls the package's
Objective-C++/Metal UMAP optimizer directly. It does not use Python,
`reticulate`, Torch, or MLX, and it does not fall back to CPU while reporting a
GPU backend.

The Metal UMAP source keeps three optimizer kernels for diagnostics:
`scheduled`, `atomic_delta`, and `torchdr_row_negatives`. The default is
`scheduled`, and benchmark scripts force `scheduled` so experimental modes are
not mixed into reference comparisons.

Native openTSNE Metal is conditional: it is available only when the package is
compiled with the `knn_tsne_opentsne_metal_cpp` symbol. In that build,
`negative_gradient_method = "auto"` resolves to the package-native Metal
FFT-grid path. If the symbol is absent, `opentsne_knn(..., backend = "metal")`
fails clearly instead of running on CPU and reporting Metal.

For CPU runs, `negative_gradient_method = "auto"` resolves to
`negative_gradient_method = "fft"`, a native FIt-SNE-style grid approximation
for the negative gradient. This is useful for large two-dimensional embeddings
from a supplied KNN graph. The older Barnes-Hut route has been removed.
Native Metal FFT openTSNE uses Objective-C++/Metal kernels when compiled.
Native CUDA FFT openTSNE still fails clearly until the CUDA port is added.

```{r, eval = fastEmbedR:::embedding_metal_available_cpp()}
metal_layout <- umap_knn(
  knn,
  backend = "metal",
  seed = 4
)

attr(metal_layout, "fastEmbedR_config")[c("backend", "optimizer", "graph_storage")]
```

## Landmark Workflow

`landmark_tsne()` embeds a subset with `opentsne()` and places the remaining
rows with the openTSNE-style fixed-reference transform.

When native openTSNE Metal is compiled, `backend = "metal"` can use fused Metal
projection kernels for query-to-landmark KNN, interpolation, and confidence
scoring. In this restored build, request Metal landmark t-SNE only after
checking `fastEmbedR:::metal_opentsne_native_available()`. CPU runs keep the
exact multi-threaded projection KNN for small and medium data, and switch
internally to a projection-specific candidate search only for larger
projections.

```{r}
landmark_fit <- landmark_tsne(
  x,
  labels = labels,
  landmarks = 0.5,
  n_neighbors = 30,
  perplexity = 10,
  early_exaggeration_iter = 100,
  n_iter = 250,
  transform_iter = 100,
  seed = 4,
  silhouette_sample = NULL,
  preserve_sample = NULL
)

plot(landmark_fit)
```

## Transform New Points

```{r}
reference <- x[1:100, , drop = FALSE]
query <- x[101:110, , drop = FALSE]

ref_fit <- opentsne(
  reference,
  n_neighbors = 30,
  perplexity = 10,
  early_exaggeration_iter = 100,
  n_iter = 250,
  standardize = FALSE,
  seed = 5,
  silhouette_sample = NULL,
  preserve_sample = NULL
)

query_knn <- nn(reference, query, k = 15)
query_layout <- transform_tsne(
  ref_fit$layout,
  knn = query_knn,
  perplexity = 5,
  n_iter = 100,
  seed = 5
)

plot(ref_fit$layout, pch = 21, bg = labels[1:100])
points(query_layout, pch = 24, bg = labels[101:110], cex = 1.2)
```

## Backend Reporting

`backend_info()` reports optional CUDA, Metal, FAISS, and cuVS availability.
Requested unavailable GPU/accelerator backends fail clearly rather than being
silently replaced by CPU and reported as GPU.

```{r}
backend_info()
```
