---
title: "EXPERIMENTAL: out-of-core analysis with fastEmbedR"
author: "Stefano Cacciatore"
output: rmarkdown::html_vignette
vignette: >
    %\VignetteIndexEntry{EXPERIMENTAL: out-of-core analysis with fastEmbedR}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Scope

This interface is **EXPERIMENTAL**. The current implementation reads a
row-major float32 file in bounded blocks, performs streamed covariance PCA
on CPU or CUDA, or randomized streamed PCA on CPU or CUDA for wide
matrices, and writes float32 PCA scores to another file. CUDA uses
the GPU for covariance or sketch blocks and score projection; the
streamed mean/scale and small eigendecomposition run on CPU. It also
supports
streaming reservoir landmarks, file-backed query-to-landmark KNN, and
fixed-reference landmark projection for UMAP or t-SNE, and landmark
Walktrap, Louvain, or Leiden clustering. CPU voting supports multiple
workers; CUDA can run landmark-graph Louvain or Leiden. CUDA can
search a resident landmark reference with exact cuVS or IVF-Flat and
write the query KNN in bounded batches. Explicit
`method = "sharded_exact"` builds one bounded CUDA reference shard at a
time and merges exact neighbors into disk-backed output. It avoids
requiring the complete reference in VRAM but rereads every query for
each shard. The single-GPU route can checkpoint every 100 query blocks
and resume a matching partial search. With multiple CUDA devices,
disjoint query ranges search bounded reference shards independently
and are merged in source order; shard and merge checkpoints can resume
matching partial work.
Neither configuration is a scalable ANN index. CUDA can also
project saved UMAP KNN rows when `refinement_epochs = 0L`, or apply the
existing fixed-reference t-SNE transform to 2D layouts in bounded
batches. Results record the CUDA backend and a VRAM estimate. CUDA
landmark UMAP refinement and Metal streaming are **not implemented**.
Experimental CPU and single-device CUDA full-graph UMAP are available
from a symmetric disk-backed fuzzy graph. File-backed t-SNE consumes
a prepared affinity graph; its CUDA optimizer currently supports 2D.
The fuzzy graph supports
experimental CPU Louvain, Leiden, and partitioned Walktrap. The
Walktrap result is approximate when more than one block is needed.
By default CUDA keeps the 2D
or 3D layout on one GPU and rejects a layout beyond its VRAM budget.
Explicit managed storage can page coordinates from host memory on a
supporting GPU; it is not a disk-backed layout.
Independent CUDA KNN and
landmark-projection query ranges can use multiple devices. Out-of-core
CUDA PCA computes its decomposition on the first selected device and
can project disjoint score rows on multiple devices. This PCA route
requires file-backed input. Optional checkpointing saves the shared
decomposition and completed shard blocks for resume with the same
devices. Landmark fitting still uses one GPU. Each KNN worker builds
its own resident landmark index, or bounded reference shards when
explicit `method = "sharded_exact"` is selected.
Unsupported routes fail explicitly.

The ordinary `pca()`, `umap()`, and `tsne()` routes are unchanged. To opt
in, pass a `massive_matrix()` descriptor with `massive = "out_of_core"`
for PCA or `massive = "landmark"` for UMAP and t-SNE. UMAP also accepts
`massive = "out_of_core_graph"` with a prepared fuzzy graph and file-backed
initial coordinates; this full-graph route currently uses one CPU worker.
Experimental landmark UMAP and t-SNE require the suggested `float`
package: selected float32 rows enter the reference fit without an R
double-matrix expansion. If `float` is absent, these routes fail
explicitly; out-of-core PCA and graph workflows do not require it.
PCA also accepts `massive = "auto"`. It estimates ordinary PCA's peak
memory, caps the supplied budget using currently available host RAM, and
uses the in-memory route only below half that effective budget and when
dimensions fit an R matrix. It reports its choice. When it selects
streaming, provide a persistent `.f32` `output` path. A resident R matrix
that exceeds the budget must first be supplied as a file-backed source;
auto mode does not make a hidden full-size copy. For a small source with
an output path, request `"out_of_core"` explicitly.
`massive_cluster()` composes landmark selection, the resident landmark
graph, and file-backed weighted-vote assignment. It accepts
`massive = "landmark"` or `massive = "auto"`; auto mode uses ordinary
`graph_cluster()` when the input fits its conservative memory budget.
The landmark route is approximate and requires an output prefix. It
does not replace `graph_cluster()` or cluster the full data graph.
UMAP and t-SNE require an explicit choice between the ordinary resident
route, approximate landmark projection, and a prepared full-graph route.
They reject `massive = "auto"` because memory pressure must not change
which observations participate in optimization. Landmark mode requires
a file-backed source, a chosen landmark count, and a persistent `.f32`
output path. The full-graph route consumes a prepared file-backed graph
and initial coordinates; it does not build the graph automatically.
Metal can run the ordinary in-memory route but has no experimental
streamed embedding route. A requested unavailable backend fails.
On Linux, PCA's auto free-RAM check respects a visible cgroup limit;
on macOS it uses the system memory-pressure report. If availability
cannot be measured, PCA auto mode asks for an explicit mode.
The public `as.matrix()` inspection method remains capped at 128 MB.
`memory_limit` budgets algorithm buffers and any resident R-matrix input;
it is not a process-memory limit. `resources$peak_ram_bytes` estimates
the largest workflow stage, including the landmark fit, KNN, and projection
for landmark embeddings. `resources$peak_vram_bytes` similarly covers the
reported CUDA stages. These are estimates, not measured peaks or guaranteed
ceilings. The R runtime, linked libraries, and CUDA context can raise actual
resident set size (RSS) substantially.

# File layout

- `.f32`: raw, little-endian, row-major float32 values. Supply `nrow` and
  `ncol` to `massive_matrix()`.
- `.fbin`: two little-endian unsigned 32-bit dimensions, followed by the
  same float32 values. Dimensions come from the header.

The file must have exactly the expected byte count. `access = "stream"`
performs sequential reads; `access = "mmap"` uses a read-only mapping on
POSIX systems. Mapping does not itself load the entire file into RAM.
`massive_coarse_postings()` can group rows by supplied centers into
physical posting files with original row IDs and 64-bit offsets. It
builds storage only: no ANN search or recall guarantee is implied.
After a successful build, `massive_open_postings()` loads the saved
manifest and centers for use in a later R session without rebuilding.
`checkpoint = TRUE` saves completed write-pass blocks. After an
interruption, rerun with `resume = TRUE` and unchanged source, centers,
output prefix, and controls. An interrupted count pass starts again;
recovery during final file renames or manifest writing is not supported.

# Example

```{r}
data <- as.matrix(iris[, 1:4])
input <- tempfile(fileext = ".f32")
scores_file <- tempfile(fileext = ".f32")
connection <- file(input, "wb")
writeBin(as.vector(t(data)), connection, size = 4L, endian = "little")
close(connection)

source <- fastEmbedR::massive_matrix(
    input, nrow = nrow(data), ncol = ncol(data)
)
posting_prefix <- tempfile()
centers <- data[c(1L, 51L, 101L), , drop = FALSE]
postings <- fastEmbedR::massive_coarse_postings(
    source, centers,
    output = posting_prefix, chunk_rows = 32L,
    n.cores = 2L, memory_limit = "128MB"
)
reopened <- fastEmbedR::massive_open_postings(posting_prefix)
postings <- reopened$postings
centers <- reopened$centers
postings$counts
fastEmbedR::massive_read_posting(postings, 1L, n = 2L)$row_id
neighbors <- fastEmbedR::massive_search_postings(
    source, postings, centers, n = 6L, k = 3L,
    nprobe = postings$nlist, exclude_self = TRUE,
    memory_limit = "128MB"
)
neighbors$indices
pilot <- fastEmbedR::massive_posting_recall_pilot(
    source, postings, centers, k = 3L,
    nprobe = postings$nlist, sample_rows = 6L,
    memory_limit = "128MB"
)
pilot[c("observed_recall", "minimum_row_recall",
    "sampled_rows_below_target", "sample_target_met")]
stopifnot(pilot$sample_target_met)
posting_graph <- fastEmbedR::massive_full_knn_graph(
    source, k = 3L, output = tempfile(),
    method = "coarse_postings", postings = postings,
    centers = centers, nprobe = postings$nlist,
    chunk_rows = 32L, reference_chunk_rows = 32L,
    memory_limit = "128MB", audit_rows = 5L,
    query_order = "posting"
)
posting_graph$observed_recall
posting_graph[c("minimum_row_recall", "sampled_rows_below_target",
    "sample_target_met")]
fit <- fastEmbedR::pca(
    source, ncomp = 2L, backend = "cpu", n.cores = 2L,
    massive = "out_of_core", output = scores_file,
    chunk_rows = 32L, memory_limit = "64MB",
    checkpoint = TRUE
)
fit <- fastEmbedR::massive_open_pca(scores_file)
fit$method
head(fit$scores)

landmark_file <- tempfile(fileext = ".f32")
neighbors_prefix <- tempfile()
selection <- fastEmbedR::massive_select_landmarks(
    source, landmarks = 30L, output = landmark_file,
    seed = 42L, chunk_rows = 32L
)
neighbors <- fastEmbedR::massive_landmark_knn(
    source, selection, k = 5L, output = neighbors_prefix,
    n.cores = 2L, chunk_rows = 32L, memory_limit = "256MB",
    checkpoint = TRUE
)
neighbors[c("nrow", "ncol", "method", "indices_path")]
audited <- fastEmbedR::massive_audit_landmark_knn(
    neighbors, source, selection, sample_rows = 6L,
    memory_limit = "256MB"
)
audited[c("observed_recall", "minimum_row_recall",
    "sampled_rows_below_target", "sample_target_met",
    "audit_sample_rows")]

stream_prefix <- tempfile()
stream_neighbors <- fastEmbedR::massive_landmark_knn(
    source, selection, k = 5L, output = stream_prefix,
    backend = "cpu", method = "stream_exact", n.cores = 2L,
    chunk_rows = 16L, reference_chunk_rows = 12L,
    memory_limit = "256MB"
)
stopifnot(identical(
    head(stream_neighbors)$indices, head(neighbors)$indices
))

reference <- fastEmbedR::massive_read_rows(selection$data, n = 30L)
reference_fit <- fastEmbedR::umap(
    reference, n_neighbors = 5L, backend = "cpu", n.cores = 2L
)
layout_file <- tempfile(fileext = ".f32")
projected <- fastEmbedR::massive_project_landmarks(
    reference_fit, neighbors, output = layout_file,
    selection = selection,
    n.cores = 2L, chunk_rows = 32L, refinement_epochs = 0L,
    checkpoint = TRUE
)
head(projected$layout)

local_file <- tempfile(fileext = ".f32")
local_projected <- fastEmbedR::massive_project_landmarks(
    reference_fit, neighbors, output = local_file,
    selection = selection, source = source,
    local_refine = TRUE, local_neighbors = 3L,
    overlap_rows = 4L, refinement_epochs = 2L,
    n.cores = 2L, chunk_rows = 32L, memory_limit = "256MB"
)
head(local_projected$layout)

umap_file <- tempfile(fileext = ".f32")
streamed_umap <- fastEmbedR::umap(
    source, massive = "landmark", landmarks = 30L,
    landmark_method = "random",
    n_neighbors = 5L, transform_k = 5L, backend = "cpu",
    output = umap_file, chunk_rows = 32L,
    memory_limit = "256MB", refinement_epochs = 0L,
    checkpoint = TRUE
)
head(streamed_umap$layout)

tsne_file <- tempfile(fileext = ".f32")
streamed_tsne <- fastEmbedR::tsne(
    source, massive = "landmark", landmarks = 30L,
    perplexity = 5, early_exaggeration_iter = 5L, n_iter = 5L,
    transform_perplexity = 3, transform_iter = 5L,
    backend = "cpu", output = tsne_file, chunk_rows = 32L,
    memory_limit = "256MB", checkpoint = TRUE
)
head(streamed_tsne$layout)

reference_graph <- fastEmbedR::knn_graph(
    reference, k = 5L, backend = "cpu"
)
cluster_prefix <- tempfile()
clusters <- fastEmbedR::massive_cluster_landmarks(
    reference_graph, neighbors, output = cluster_prefix,
    method = "leiden", selection = selection,
    n.cores = 2L, chunk_rows = 32L
)
head(clusters)

cluster_workflow <- fastEmbedR::massive_cluster(
    source, method = "leiden", massive = "landmark",
    landmarks = 30L, k = 5L, backend = "cpu",
    n.cores = 2L, chunk_rows = 32L, memory_limit = "256MB",
    output = tempfile()
)
head(cluster_workflow)
unlink(c(input, scores_file, paste0(scores_file, ".manifest.rds"),
    landmark_file, layout_file, local_file,
    postings$features$path, postings$ids_path, postings$offsets_path,
    postings$manifest_path,
    neighbors$indices_path, neighbors$distances_path,
    stream_neighbors$indices_path, stream_neighbors$distances_path,
    clusters$membership_path, clusters$confidence_path))
unlink(c(cluster_workflow$selection$data$path,
    cluster_workflow$knn$indices_path,
    cluster_workflow$knn$distances_path,
    cluster_workflow$membership_path,
    cluster_workflow$confidence_path))
for (result in list(streamed_umap, streamed_tsne)) {
    unlink(c(result$layout$path, result$selection$data$path,
        result$reference_fit_path, result$graph$indices_path,
        result$graph$distances_path))
}
```

The optional posting-order full-graph route groups queries by physical
posting before searching, then externally restores original row order in
the saved graph. It can reduce repeated posting reads when RAM limits
force small query batches, at the cost of temporary disk space and a
reorder pass. The default `query_order = "original"` avoids that pass.
Both routes are CPU-only; neither is a general ANN recall guarantee.
With `checkpoint = TRUE`, a completed grouped search can be reused if
the reorder pass fails; rerun with `resume = TRUE` and identical inputs.

For CUDA landmark Louvain or Leiden, `devices = c(0L, 1L)` assigns
disjoint query-to-landmark searches to two GPUs. The first device builds
and clusters the resident landmark graph; CPU workers still vote labels
from file-backed neighbor rows. Each selected GPU must hold the reference
index. A multi-GPU landmark run requires an explicitly installed CUDA
backend and fails if either device is unavailable; it never switches to CPU.

`fit$scores` is a file descriptor, not a full R matrix. Use
`massive_read_rows()` to inspect a bounded range. `as.matrix()` refuses
materialization above 128 MB. The output path must not already exist and
the destination disk must have sufficient free space. A completed PCA saves
its loadings and score-file identity beside the `.f32` file; reopen it in a
later R session with `massive_open_pca(output)`. Reopening rejects a changed
or incomplete score file, so downstream graph jobs do not silently consume
stale coordinates.

With `checkpoint = TRUE`, covariance PCA saves completed mean and
covariance blocks;
wide randomized PCA saves completed column moments. Both then save the
fitted decomposition and each flushed score batch. Resume with the same
settings and `resume = TRUE`; it verifies the saved stage before reuse.
Covariance work since the last saved block repeats after interruption.
An interrupted wide moments pass or sketch action restarts that pass.
Projection leaves `output.part` and `output.checkpoint.rds` until complete.
For public landmark `umap()` and `tsne()`, `checkpoint = TRUE` also saves
selection, reference-fit, and completed-graph identities. After an
interruption, repeat the same call with `resume = TRUE`; changed inputs or
settings fail rather than silently repeating completed stages. Keep all
sidecar files and partial outputs together until the run completes.

The streamed covariance route handles up to 1,024 columns on CPU and
2,048 columns on CUDA. Wider matrices use a seeded randomized feature
sketch with two streamed
covariance-action passes. Its accuracy depends on the spectrum and sketch
size; it is an approximation, and the result reports the selected method.
Wider CUDA matrices use a streamed GPU randomized sketch. Metal streaming
is not implemented. The CUDA covariance route keeps reusable device buffers
for each covariance or projection pass; it still reads each block into
host RAM before transfer. The
eigendecomposition itself is CPU-based and requires only the small
covariance matrix. File-backed PCA scores can be passed directly to experimental
`umap(massive = "landmark")` or `tsne(massive = "landmark")` without
materializing all scores. This is a landmark approximation, not full
out-of-core optimization of either objective or the complete graph.
The unified calls save the landmark sample, reference fit, neighbor
files, and output layout beside the requested `.f32` file. With
`checkpoint = TRUE`, both calls save completed workflow stages and
projection batches. `massive_cluster()` likewise saves the selected
landmarks, resident reference graph, query-to-landmark KNN, and completed
label batches. To continue after interruption, preserve the output and
sidecars and repeat the same call with `resume = TRUE`. The resident
reference graph is serialized in the checkpoint, so opt-in checkpointing
requires additional disk space. A crash between finalizing a stage file
and committing its workflow checkpoint still needs manual inspection;
the package refuses to overwrite that uncommitted artifact.

`massive_select_landmarks()` is a separate experimental preparation step.
It selects row identifiers with a seeded reservoir algorithm, then streams
the source once to write those rows to a new `.f32` file in source order.
For fast random access, `landmark_method = "random"` selects in time
proportional to the landmark count and reads only the sampled source
rows. The option also works in experimental landmark `umap()`, `tsne()`,
and `massive_cluster()` calls. It may be slower on storage with expensive
random reads.
The returned `data` is file-backed; it is not yet a fitted UMAP or t-SNE
landmark model. `massive_landmark_knn()` normally keeps the landmark
reference and its search index in RAM or VRAM, searches the input in
bounded batches, and writes row-major, one-based `uint32` landmark IDs
and float32 distances.
The optional audit compares selected rows with a streamed exact CPU
reference. Its measured recall describes only those rows; it does not
guarantee the requested target for the remaining observations.
File-backed HNSW and CUDA references enter the native index as float32;
`reference_storage` reports this as `"native_float32"`. Small CPU exact
search still uses a resident R double reference.
For a reference that does not fit in RAM, set `method = "stream_exact"`
on CPU and optionally set `reference_chunk_rows`. This scans bounded
query and reference blocks without loading the reference. It is exact
but quadratic in the number of queries and landmarks, so it is a
correctness and memory-limited route rather than a scalable ANN index.
It does not silently select CUDA or replace the default method.
For fewer than 5,000 landmarks, the default is exact search; otherwise a
persistent HNSW index is built once. CUDA uses exact cuVS below 100,000
landmarks and IVF-Flat above that threshold. HNSW's 0.99 recall target
is not an observed-recall guarantee; IVF-Flat reports a first-batch
pilot recall only. The returned KNN files are not a full fuzzy UMAP
graph.

An external full-data KNN producer can write the same pair of files with
exactly `k` distinct, non-self neighbors for every row. Import them without
copying the files:

```{r eval=FALSE}
graph <- fastEmbedR::massive_knn_graph(
    "deep1b_knn", n_vertices = 1e9, k = 30, access = "mmap"
)
edges <- fastEmbedR::massive_read_graph_edges(graph, first = 1, n = 100)
partitions <- fastEmbedR::massive_graph_partitions(graph,
    max_bytes = 64e6)
first_partition <- fastEmbedR::massive_read_graph_partition(
    partitions, 1)
```

For a small or moderate file-backed matrix, an exact CPU graph can be
built without loading the whole matrix:

```{r eval=FALSE}
graph <- fastEmbedR::massive_full_knn_graph(
    source, k = 30L, output = "full_knn",
    n.cores = 8L, chunk_rows = 256L,
    reference_chunk_rows = 8192L,
    memory_limit = "8GB", checkpoint = TRUE,
    checkpoint_every = 100L
)
```

This route streams query and reference blocks, writes the same two graph
files, and can resume completed query blocks. It is quadratic in row
count and is a correctness baseline, not an ANN route suitable for
million- or billion-row production graphs. CUDA exact search is also
available: it builds one cuVS index per bounded reference shard and
merges the neighbor candidates into the same file-backed output.

```{r eval=FALSE}
cuda_graph <- fastEmbedR::massive_full_knn_graph(
    source, k = 30L, output = "full_knn_cuda",
    backend = "cuda", chunk_rows = 256L,
    reference_chunk_rows = 8192L,
    memory_limit = "8GB"
)
```

The exact CUDA route is likewise quadratic, requires `k <= 255`, and supports
opt-in checkpoints. By default a checkpoint is written every 100 query
blocks and at every shard boundary; uncommitted blocks are replayed on
resume. Use the same `checkpoint_every` value when resuming. An explicit
`method = "ivf_sharded"` uses native cuVS IVF-Flat within each reference
shard. It is approximate: the 0.99 target is requested, not guaranteed
for the merged graph. Set `audit_rows` to measure sampled recall. Both
CUDA methods fail when native cuVS is unavailable; neither falls back
to CPU. Repeated query scans still limit scaling. The saved graph can
feed the CPU full-graph
UMAP route below. The derived fuzzy graph can also feed experimental
Louvain clustering as described below.

For an explicitly approximate CPU route, `method = "hnsw_sharded"`
builds one bounded native HNSW index per reference shard and merges its
query results into the same file-backed graph format. It reports the
index-build and query/merge times separately. Set `audit_rows` to compare
evenly spaced query rows against streamed exact neighbors and record mean
and minimum row recall. Each audited row scans the full source, so budget
this separately. Opt-in checkpoints use the same commit interval.
Resume rebuilds the shard index and compares a pilot query before
continuing. Repeated full query scans can make the route I/O-intensive,
and billion-row performance is not established.

Import scans all edges for malformed IDs, distances, self edges, and
duplicates using bounded sequential reads, even when later edge reads use
mmap. A bounded read returns directed edges with distances, not UMAP
fuzzy weights. The partition plan is constant-size even for a billion-row
graph and caps each returned R edge list. The sharded HNSW route avoids a
full-matrix index but is not yet a scale-validated full-data ANN builder.
Only the symmetric fuzzy graph produced below can feed the experimental
full-graph UMAP optimizer. It can also be scanned to measure modularity
or clustered with file-backed Louvain. Contracted levels remain
file-backed while too large for a conservative resident-memory check.
The storage API permits no more than
2^31 - 1 vertices and uses exact 64-bit byte offsets internally.
Directed UMAP memberships can be calculated from a distance graph without
loading all edges into memory:

```{r eval=FALSE}
weighted <- fastEmbedR::massive_umap_memberships(graph, n.cores = 4L)
weighted_edges <- fastEmbedR::massive_read_graph_edges(
    weighted, first = 1, n = 100)
```

This scans the distances twice and writes `full_knn.weights.f32` beside
the existing KNN files. Independent rows can use multiple CPU workers;
graph sorting and full-graph optimization have separate threading limits.
An explicit CUDA request fails. Set `checkpoint = TRUE` to save completed
mean-scan and weight blocks; after interruption, rerun with
`resume = TRUE`.

The directed memberships can be symmetrized with the UMAP fuzzy union
without holding all edges in memory:

```{r eval=FALSE}
fuzzy <- fastEmbedR::massive_umap_fuzzy_graph(
    weighted, output = "full_knn_fuzzy", memory_limit = "8GB",
    checkpoint = TRUE, checkpoint_every = 100L
)
fuzzy_edges <- fastEmbedR::massive_read_graph_edges(
    fuzzy, first = 1, n = 100)
```

This CPU route externally sorts edges into a variable-degree CSR graph.
It checks temporary disk space before starting. If interrupted, repeat
the same call with `resume = TRUE` and unchanged inputs and controls.
Completed pair-sort runs are reused; the final merge and CSR write repeat.
Checkpointed sorting retains more temporary data than an uninterrupted
sort. A failed final output commit may need manual inspection. It writes a
`full_knn_fuzzy.fuzzy.rds` manifest containing graph hashes. After a
restart, reload the same graph without recomputing neighbors or fuzzy
memberships:

```{r eval=FALSE}
fuzzy <- fastEmbedR::massive_fuzzy_graph(
    "full_knn_fuzzy", n_vertices = graph$n_vertices
)
```

Reloading scans and hashes the CSR files. A generic `massive_csr_graph()`
import does not certify a graph as UMAP fuzzy input.

The fuzzy graph can supply a multilevel Louvain fit:

```{r eval=FALSE}
communities <- fastEmbedR::massive_cluster(
    fuzzy, method = "louvain", massive = "out_of_core_graph",
    output = "full_louvain.clusters.u32",
    memory_limit = "8GB"
)
head(communities)
```

This experimental CPU-only route memory-maps community state, scans
graph blocks, and externally contracts edges. Contracted levels remain
file-backed until a conservative memory check permits resident Louvain.
Use `massive_louvain_level()` to request only the first local-moving level.
The same fuzzy graph also supports file-backed Leiden:

```{r eval=FALSE}
leiden <- fastEmbedR::massive_cluster(
    fuzzy, method = "leiden", massive = "out_of_core_graph",
    output = "full_leiden.clusters.u32",
    memory_limit = "8GB", checkpoint = TRUE
)
head(leiden)
```

Leiden keeps every graph level file-backed and carries each refined
partition into the next local-moving level. Partitioned Walktrap reuses
the native exact walk kernel within bounded induced blocks, contracts
their communities, and repeats on the contracted graph:

```{r eval=FALSE}
walktrap <- fastEmbedR::massive_cluster(
    fuzzy, method = "walktrap", massive = "out_of_core_graph",
    output = "full_walktrap.clusters.u32", steps = 4L,
    chunk_rows = 4000L, memory_limit = "8GB"
)
head(walktrap)
walktrap$approximate
```

Cross-block walks are absent from the first level. This CPU-only route
is sensitive to vertex order, is not mathematically identical to
whole-graph Walktrap, and cannot yet
checkpoint or resume. If a level cannot contract, it fails and retains
partial labels for inspection. Full-graph Leiden uses one CPU worker and does
support restart from completed graph levels with `resume = TRUE` and
unchanged inputs and controls.
Mapped pages may count toward resident memory; `memory_limit` is not a
peak-RSS guarantee. For stage-boundary restart, use `checkpoint = TRUE`
in a full-graph Louvain or Leiden `massive_cluster()` call. Repeat with
`resume = TRUE` and unchanged inputs and controls. Partial
native scans and contraction work files cannot yet resume; Leiden retains
its completed contracted levels, while Louvain checkpoints only its
first level and first contraction.

A distance-valued full KNN graph can also be converted to file-backed
t-SNE affinities with the same native float32 probability policy:

```{r eval=FALSE}
affinities <- fastEmbedR::massive_tsne_affinities(
    graph, perplexity = 30, output = "full_knn_tsne",
    memory_limit = "8GB", checkpoint = TRUE
)
```

The affinity graph can feed the experimental full-data FFT optimizer on
CPU or CUDA. CUDA currently accepts two-dimensional coordinates only.
Supply pre-scaled PCA or spectral coordinates in a row-major float32 file,
with the same row count and two or three columns:

```{r eval=FALSE}
tsne_initial <- fastEmbedR::massive_matrix(
    "tsne_initial.f32", nrow = affinities$n_vertices, ncol = 2L
)
tsne_layout <- fastEmbedR::tsne(
    affinities, perplexity = 30, Y_init = tsne_initial,
    massive = "out_of_core_graph", backend = "cuda",
    output = "full_tsne.f32", memory_limit = "32GB",
    checkpoint = TRUE
)
```

The optimizer streams the affinity graph on every iteration. CPU retains
float32 optimizer state in RAM; CUDA retains it on one GPU and fails when
the layout and fast Fourier transform workspace exceed free GPU memory.
The graph remains file-backed in both routes. Large-graph runtime and
quality still need measurement.
An interrupted fit can continue with the same call and `resume = TRUE`.
Snapshots retain coordinates, momentum updates, and adaptive gains;
uncommitted partial snapshots require inspection before resume.
The optional `n.cores` argument in affinity construction
parallelizes conditional-probability rows,
not the external graph sort; it did not improve total time in a
million-row compact-support check. An interrupted pair sort can resume
with `resume = TRUE` and identical inputs and controls.

To fit an experimental UMAP layout from the fuzzy graph, supply a
file-backed 2D or 3D initialization with one row per observation:

```{r eval=FALSE}
layout <- fastEmbedR::umap(
    fuzzy, massive = "out_of_core_graph", init = fit$scores,
    output = "full_umap.f32", memory_limit = "8GB",
    n_components = 2L, backend = "cpu", n.cores = 1L,
    layout_storage = "mmap"
)
```

For checkpointing and optimizer controls, use the same graph with the
lower-level function and a different output file:

```{r eval=FALSE}
layout <- fastEmbedR::massive_umap_optimize(
    fuzzy, init = fit$scores,
    output = "full_umap_checkpointed.f32", memory_limit = "8GB",
    checkpoint = TRUE, checkpoint_every = 10L
)
```

The optimizer streams the graph for each epoch and currently uses one CPU
worker. The default retains float32 coordinates in RAM. Explicit
`layout_storage = "mmap"` keeps coordinates in a writable file instead;
this requires a POSIX system, sufficient disk space, and can be slower
because edge updates access coordinates irregularly. The OS may cache
mapped pages in RAM, so `memory_limit` is not a hard process-RSS cap.
An interrupted fit in either storage mode can be restarted with the same
inputs and controls plus `resume = TRUE`; only complete epoch snapshots
are reused. Checkpointing writes a full snapshot at each interval and
needs space for two coordinate files in RAM mode or three in mapped mode.
CUDA full-graph optimization streams bounded edge batches from the same
prepared fuzzy graph. The default keeps 2D or 3D coordinates on one GPU
and requires the layout to fit device memory. Opt in to
`layout_storage = "managed"` to use CUDA unified memory on a GPU that
supports concurrent managed access. Managed coordinates count against
the host-memory budget and may page slowly if they exceed VRAM. This
does not provide disk-backed coordinates. CUDA rejects `"mmap"`.
CUDA can use the same completed-epoch checkpoint controls;
each snapshot downloads the full layout and needs additional disk space.
Resume requires the same backend, graph, initialization, and controls.
It does not silently use CPU optimization.
The reported `device_buffer_bytes` counts CUDA device allocations;
`managed_layout_bytes` reports the separate coordinate allocation.
Neither includes CUDA runtime or allocator overhead. Chunk boundaries
change the order
of asynchronous edge updates, so streamed and resident coordinates need
not match exactly even with identical graph, seed, and initialization.
CUDA graph sorting is not implemented.
An external producer may also supply
`.indices.u32` and `.weights.f32` for `massive_weighted_graph()`.
Weight import checks values in [0, 1] and returns a `weight` field, but
does not verify graph symmetry.

For a variable-degree weighted graph, use a CSR file triple:

```{r eval=FALSE}
graph <- fastEmbedR::massive_csr_graph(
    "deep1b_csr", n_vertices = 1e9, access = "mmap"
)
partitions <- fastEmbedR::massive_graph_partitions(graph)
edges <- fastEmbedR::massive_read_graph_partition(partitions, 1)
```

The files are `.offsets.u64`, `.indices.u32`, and `.weights.f32`.
Offsets are zero-based positions in the edge files; neighbor IDs are
one-based and strictly increasing within each row. Validation reads every
edge sequentially, and the partition plan uses the maximum row degree to
bound returned R edge lists. Import does not calculate fuzzy memberships
or prove that an externally supplied graph is symmetric. Graphs returned
by `massive_umap_fuzzy_graph()` have been constructed with fuzzy union.

For long KNN searches, `checkpoint = TRUE` saves completed rows in a
`.checkpoint.rds` sidecar beside the two `.part` files. After an
interruption, rerun the same call with `resume = TRUE`. The reference
index is rebuilt, and its original first-batch output must match the
saved pilot before either output file is extended. The file-backed
inputs must retain their paths, sizes, and modification times. This
pilot detects changed search behavior but is not a proof that every
approximate-index entry is identical after a restart. For exact search,
it also validates the original input batch. A completed run removes
the sidecar and finalizes both output files. With `backend = "cuda"` and
`devices = c(0L, 1L)`, KNN searches disjoint query ranges and merges
their row-major output files. Resident exact/IVF search requires the
landmark reference and index to fit on each device. Explicit
`method = "sharded_exact"` instead builds bounded reference shards on
each device, at the cost of a complete query pass per shard. Multi-GPU
KNN currently rejects `checkpoint` and `resume`; single-device KNN
retains the checkpoint behavior above.

`massive_project_landmarks()` fits no new reference embedding: fit
`umap()` or `tsne()` on the selected landmark matrix first, then project
query rows from the saved KNN graph. The output is a file-backed query
layout. Landmark coordinates stay fixed during projection. When the
queries include the source rows, `selection` replaces projected landmark
rows with their fitted reference coordinates. With zero UMAP refinement
epochs it is weighted
KNN projection only. Positive refinement epochs or t-SNE transform
iterations reuse the package's fixed-reference optimizers. Stochastic
refinement is batch-dependent, so retain the chunk size and seed when
comparing results.

A high-level landmark fit can also reuse a previous fit's saved neighbor
search. Pass the previous result as `nn` with the same source, landmark
count, projection neighbor count, and backend. The second call fits its
own reference embedding but does not select landmarks or search the
queries again. It fails rather than rebuilding if the source, landmark
reference, saved neighbor files, or requested neighbor count differs.

```{r massive-reuse, eval=FALSE}
first <- fastEmbedR::umap(source, massive = "landmark",
    landmarks = 10000L, n_neighbors = 15L, transform_k = 15L,
    output = "umap.f32", backend = "cpu")
second <- fastEmbedR::tsne(source, massive = "landmark",
    landmarks = 10000L, nn = first, perplexity = 10,
    transform_k = 15L, transform_perplexity = 5,
    output = "tsne.f32", backend = "cpu")
stopifnot(second$graph_reused)
```

The same saved selection and query KNN can feed landmark Leiden,
Louvain, or Walktrap clustering without another all-row search. The
resident graph among landmarks is still constructed for clustering;
cluster labels and confidence are written as file-backed outputs.

```{r massive-cluster-reuse, eval=FALSE}
communities <- fastEmbedR::massive_cluster(source,
    method = "leiden", massive = "landmark", nn = first,
    k = 15L, output = "communities", backend = "cpu")
stopifnot(communities$graph_reused)
```

The reused result must retain its original source, landmark, and KNN
files. A different source, backend, or neighbor count fails explicitly.

`local_refine = TRUE` is an opt-in **EXPERIMENTAL** CPU UMAP variant.
Pass the original file-backed query `source`, the exact saved KNN graph
created from it, positive `refinement_epochs`, and a bounded
`local_neighbors` count. It reconstructs query-to-query neighbors inside
each chunk plus `overlap_rows` on both sides, then updates only the core
query rows while preserving landmark coordinates. The KNN result records
the source file identity, and mismatches fail before projection. A
conservative memory preflight can reduce the effective chunk size.
These local windows do not constitute a full-data UMAP graph. Boundary
effects remain possible, and quality improvement is dataset-dependent.
The extra KNN construction makes this route slower than anchor-only
projection on some datasets; CUDA requests fail without CPU fallback.

Set `checkpoint = TRUE` to save completed projection batches in a
`.checkpoint.rds` sidecar while writing the `.part` output. If the
process stops, rerun the same call with `resume = TRUE`. Resume requires
the same layout, KNN files, settings, and effective chunk size; changed
inputs fail rather than silently mixing results. The sidecar and `.part`
file are removed or finalized after success. File paths, sizes, and
modification times are checked, but this is not a cryptographic audit of
the complete KNN files. Do not edit those files during a run.
After completion, `massive_open_projection(output)` reopens the saved
coordinates without loading them. Its local manifest requires the same
coordinate and KNN paths, sizes, and modification times; it is not a
portable content hash.

On a CUDA-capable build, zero-epoch UMAP projection and 2D or 3D t-SNE
fixed-reference transformation can use `backend = "cuda", n.cores = 1L`.
t-SNE requires at most 128 saved neighbors. The landmark layout and each
bounded KNN batch are uploaded to the GPU; the entire query graph and
output never need to fit in VRAM. The batch size is capped from free
VRAM, and the native transform checks its workspace before each batch.
CUDA zero-epoch UMAP retains the reference layout and query buffers
when the reference transfer exceeds the planned query-batch transfers;
otherwise it uses the existing per-batch CUDA route. The chosen
`projection_engine` is reported in `resources`. CUDA t-SNE still
allocates per batch. Neither route is guaranteed to outperform CPU.

`devices = c(0L, 1L)` also opts into experimental multi-GPU query
projection on a machine exposing both CUDA devices. The reference fit
uses the first selected device. KNN searches disjoint query ranges, then
each device projects its own range from the merged KNN graph. Each
process writes a shard, which is merged in input-row order using a
bounded buffer. The returned `gpu_devices`, per-device resource
estimates, and `shard_paths` document projection execution. This route
requires additional disk space for KNN and projection shards alongside
their final outputs. Projection supports `checkpoint = TRUE` and
`resume = TRUE`: completed shards are reused, partial shards and an
interrupted merge resume. The input files, selected devices, and
effective batch settings must match; changed inputs fail explicitly.
Checkpoint identity uses file paths, sizes, and modification times, not
cryptographic hashes of complete KNN files. Reused shards report unknown
peak VRAM rather than a fabricated estimate. This route does not
distribute reference fitting or landmark optimization, and it does not
support local refinement. An unavailable device fails without CPU
fallback.

`massive_cluster_landmarks()` reuses the ordinary `knn_graph()` and
`graph_cluster()` implementations on the resident landmark subset. It
then reads the saved query-to-landmark KNN graph in batches and assigns
each query by inverse-distance weighted community vote. The confidence
file holds the winning community's fraction of vote weight. This is a
landmark approximation, not full-graph community detection. The reference
graph must fit in RAM; query labels and confidence remain file-backed.
The saved KNN graph may have been built on CPU or CUDA. Landmark-graph
Louvain and Leiden can use `backend = "cuda"`, while Walktrap remains
CPU-only. Query community voting remains on CPU in either case. The
result reports `backend`, `assignment_backend`, and `graph_backend`.
Conservative RAM and, for CUDA, VRAM estimates include the resident
landmark graph before clustering starts.
With `checkpoint = TRUE`, the landmark communities and completed query
rows are saved beside the two partial output files. Rerunning the same
call with `resume = TRUE` reuses the saved communities rather than
clustering the landmark graph again. `checkpoint_every` controls how
many query batches may need to be repeated after an interruption.
Changing the graph, KNN inputs, method, or settings rejects the resume.
